---
title: "Feelings_initial"
output: html_document
date: "2025-04-13"
---

```{r}
feelings_initial <- load("/Users/emma/Desktop/RA/Silvia_Stanford/Feelings_initial_Stanford/feelings_initial.RData")
ls()
```

```{r}
summary(feelings_initial)
```

```{r}
str(dat)
```

```{r}
head(Iaro_wide)
```

```{r}
head(Ineg_wide)
```

```{r}
head(Ipos_wide)
```


## Descriptive statistics

```{r}
summary(dat[, c("Ineg", "Ipos", "Iaro")])
```

```{r}
# identify NAs
colSums(is.na(dat))
```

There are no NAs in the dataset.

```{r}
# identify outliers using z-score

# Calculate Z-scores for Ineg, Ipos, and Iaro
dat$z_Ineg <- scale(dat$Ineg)
dat$z_Ipos <- scale(dat$Ipos)
dat$z_Iaro <- scale(dat$Iaro)

# Identify outliers (Z-score > 3 or < -3)
outliers_Ineg <- dat[abs(dat$z_Ineg) > 3, ]
outliers_Ineg
outliers_Ipos <- dat[abs(dat$z_Ipos) > 3, ]
outliers_Ipos
outliers_Iaro <- dat[abs(dat$z_Iaro) > 3, ]
outliers_Iaro

```

There are no outliers.

## Mixed-effects model: analyze data with repeated measures

- Each participant has multiple trials, so the trials within a participant are likely correlated
- Data is nested
- Each participant may have their own baseline level of emotional responses

- fixed effects (trial.val, sex, age, ethn) explain the variation between individuals
- random effects (1|subj) explain the correlation of repeated measures within individuals

```{r}
library(lme4)

# Mixed-effects model for predicting Ineg
model_ineg <- lmer(Ineg ~ trial.val + sex + age + ethn + (1|subj), data = dat)
summary(model_ineg)

```

- Random effects: each participant has a different baseline emotional response
  - `(1|subj)`: represents the random effect 
    - each participant (subj) has a different baseline deviation (intercept). 
    - This accounts for the correlation between multiple trial results from the same participant
- `REML score` (residual maximum likelihood estimate): assess the model fit
- Fixed Effects:
  - Intercept: Negative trial
  - `trial.valneu` (Neutral trial): Estimate = -4.08, t = -118.57, a very significant negative value. 
    - Compared to the baseline (negative trial), the neutral trial significantly decreases negative emotions (Ineg)
  - `trial.valpos` (Positive trial): Estimate = -4.09, t = -168.08, also significant. 
    - the positive trial also significantly decreases negative emotions compared to the negative trial
  - sexfemale: Estimate = 0.317543, t = 2.606. 
    - Females have significantly higher negative emotional responses (Ineg) compared to males
  - The effects of age and ethnicity are small and not significant


```{r}
# Mixed-effects model for predicting Ipos
model_ipos <- lmer(Ipos ~ trial.val + sex + age + ethn + (1|subj), data = dat)
summary(model_ipos)
```

- Intercept (negative trial): estimate = 0.72, t-value = 1.56. The effect of negative trial on positive emotions (Ipos) is small.
- `trial.valneu`: estimate = 0.34, t-value = 9.48. Compared to valneg, the neutral trial significantly increases positive emotions (Ipos).
- `trial.valpos`: estimate = 4.03, t-value = 160.64. Compared to valneg, the positive trial largely increases positive emotions (Ipos), and the effect is extremely significant.
- `sexfemale`: estimate = 0.20, t = 1.58. Females tend to have slightly higher positive emotional responses than males.
- `ethnAmerican Indian/Native American or Alaskan Native`: estimate = -0.94, t = -2.30. This ethnicity tends to have significantly lower positive emotional responses compared to the reference group.
- `trial.valneu` and `trial.valpos` have a correlation of 0.354, showing that the effects of neutral and positive trials are somewhat related.


```{r}
# Mixed-effects model for predicting Iaro
model_aro <- lmer(Iaro ~ trial.val + sex + age + ethn + (1|subj), data = dat)
summary(model_aro)
```

- Intercept (negative trial): estimate = 2.93, t-value = 3.84. The effect of negative trial on arousal (Iaro) is moderate.
- `trial.valneu`: estimate -2.26, t-value = -64.28. Compared to `valneg`, the neutral trial significantly decreases arousal (Iaro), which can be expected.
- `trial.valpos`: estimate = -0.30, t-value = -12.10. Compared to `valneg`, the positive trial also significantly decreases arousal (Iaro), but the effect is small.
- Other fixed effects are not significant.



