---
title: "Feelings_initial"
output: html_document
date: "2025-04-13"
---

```{r}
feelings_initial <- load("/Users/emma/Desktop/RA/Silvia_Stanford/Feelings_initial_Stanford/feelings_initial.RData")
ls()
```

```{r}
summary(feelings_initial)
```

```{r}
str(dat)
```

```{r}
head(Iaro_wide)
```

```{r}
head(Ineg_wide)
```

```{r}
head(Ipos_wide)
```


## Descriptive statistics

```{r}
summary(dat[, c("Ineg", "Ipos", "Iaro")])
```

```{r}
# identify NAs
colSums(is.na(dat))
```

There are no NAs in the dataset.

```{r}
# identify outliers using z-score

# Calculate Z-scores for Ineg, Ipos, and Iaro
dat$z_Ineg <- scale(dat$Ineg)
dat$z_Ipos <- scale(dat$Ipos)
dat$z_Iaro <- scale(dat$Iaro)

# Identify outliers (Z-score > 3 or < -3)
outliers_Ineg <- dat[abs(dat$z_Ineg) > 3, ]
outliers_Ineg
outliers_Ipos <- dat[abs(dat$z_Ipos) > 3, ]
outliers_Ipos
outliers_Iaro <- dat[abs(dat$z_Iaro) > 3, ]
outliers_Iaro

```

There are no outliers.

## Mixed-effects model: analyze data with repeated measures

- Each participant has multiple trials, so the trials within a participant are likely correlated
- Data is nested
- Each participant may have their own baseline level of emotional responses

```{r}
library(lme4)

# Mixed-effects model for predicting Ineg
model_ineg <- lmer(Ineg ~ trial.val + sex + age + ethn + (1|subj), data = dat)
summary(model_ineg)

```

```{r}
# Mixed-effects model for predicting Ipos
model_ipos <- lmer(Ipos ~ trial.val + sex + age + ethn + (1|subj), data = dat)
summary(model_ipos)
```

```{r}
# Mixed-effects model for predicting Iaro
model_aro <- lmer(Iaro ~ trial.val + sex + age + ethn + (1|subj), data = dat)
summary(model_aro)
```


- fixed effects (trial.val, sex, age, ethn) explain the variation between individuals
- random effects (1|subj) explain the correlation of repeated measures within individuals




